---
title: "Guide on Diversity Metrics"
author: "Gonzalo Velasco Mones"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: united
    code_folding: show
    code_summary: "Show code"
    code_tools: true
    code_line_numbers: true
    code_overflow: scroll
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: true
    fig_captions: true
    highlight: tango
bibliography: "https://api.citedrive.com/bib/126da26e-5aac-4b88-91f5-5b4fb1c9304b/references.bib?x=eyJpZCI6ICIxMjZkYTI2ZS01YWFjLTRiODgtOTFmNS01YjRmYjFjOTMwNGIiLCAidXNlciI6ICIxMzk2IiwgInNpZ25hdHVyZSI6ICIyY2FiZWMwMTA5ZWFjZDg2ZjVhZGZmNjcxODQ2ODQwOWNmZTYyYWI4Mzk5NzhiMmZjY2NkNGMwMGJlNDAxMDUyIn0="
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Components of diversity {.tabset .tabset-fade .tabset-pills}

Traditionally, to calculate diversity, we need a vector of relative abundances, $\mathbf{p}$, 
enclosing the abundance proportion of each species in the community. Hence, we obtain:

1. $\mathbf{p} = [p_1,\ p_2,\ \ldots,\ p_n]$, $n$ being the number of species present in our sampled community. 
2. $\sum \mathbf{p} = 1$
3. $p_i > 0\ \forall\ i \in \{1, 2, \ldots, n\}$

We can get this from any sample of species communities 
(traditionally a matrix of plots $\times$ species in vegetation ecology). 

When we include functional diversity, we need a matrix of distances $\mathbf{D}$ coming from
a matrix of untransformed trait values (even monotonic transformations like logarithms 
or square roots can alter the ordering of distances, making a pair of species close in
the natural, or arithmetic, scale farther apart in the transformed scale than a pair of species
originally distant on the natural scale). Gower distance is the most common type of 
distance to deal with qualitative and quantitative trait scales simultaneously, and
a recent modification by @debello2021 can find an appropriate weighting for traits
that belong to common organismic properties (for example to not overemphasize our
total distance from leaf traits if we include Leaf Area, SLA and LDM at the same time).

Note that the following components of diversity are better understood as concepts telling
us something about the communities than as actual metrics, implying that we must try to make
sure that the metrics used capture the variation in each of the components, but the choice
of metrics might be dependent on the research question and approach, and there might be
many that provide equally useful summaries for each of the components.

## Taxonomic Diversity

### Richness

The most common metric of diversity is of course taxonomic richness, or the number
of species found within the community. This has since early times related to the
areal distribution and to biological processes of competition or habitat suitability
(see @browne1980, @richards2012, @vandermaarel2015).

### Evenness

Evenness provides a measure of how differences in relative abundance (i.e. the different 
$p_i$ values) affect how we measure diversity. Many diversity metrics such as those coming
from information theory give a mix of richness and evenness (@whittaker1972), as they
emphasize that, given a richness number $S$, a more even community will have a higher 
diversity than a less even community. 

A common way to measure this is through  Pielou's evenness metric, although in this 
study we will directly assess how diversity changes across the rarity-abundance spectrum 
(see *Hill-Chao numbers* section).

## Functional Diversity

By the introduction of the matrix $\mathbf{D}$, the framework for functional diversity is
more complex than for taxonomic diversity @schmera2023, @debello_book2021

### Richness

In the context of functional diversity, richness aims to provide a measure of the size
of the trait space. It can be obtained as the number of functional clusters in a 
classification of the distance matrix, or as the hypervolume defined in the functional
space by our communities. 

**In this paper we use hypervolumes to quantify richness.**

### Regularity

Due to both numerical and conceptual problems with the concept of evenness in functional
space (refs), researchers have provided regularity as an alternative, but it aims to 
capture a similar idea as evenness in taxonomic diversity, i.e. how dense vs sparse
regions of the trait space alternate (i.e. a perfectly uniformly distributed 
trait space would have high regularity, while a trait space with strongly
separated sparse and dense regions would have low regularity). 

**In this paper we use the topological structure of the hypervolumes to quantify regularity.**

### Divergence

Divergence provides a measure telling us what are the expected mean dissimilarities
(i.e. the average distance value) between species. This can be obtained as the mean nearest 
neighbor distance in tree-based distance contexts or as Rao Quadratic Entropy. 

**In this paper we use Rao Quadratic Entropy, plus a correlated set of metrics
introduced later, to quantify divergence.**


# Hill-Chao numbers

Hill numbers were originally provided to offer a family of diversity metrics, unified
by a parameter $q$, that assigns different weights to common and rare species unifying 
along its domain a host of other diversity metrics. In particular, calling our diversity
of order $q$ as a function on the $\mathbf{p}$ vector $^qD(\mathbf{p})$:

1. When $q = 0$: $\ ^0D(\mathbf{p}) = S$, or species richness
2. When $q = 1$: $\ ^1D(\mathbf{p}) = \exp(H)$, or exponential of Shannon's entropy
3. When $q = 2$: $\ ^2D(\mathbf{p}) = \sum\frac{1}{p_i^2}$, or Simpson's index
4. When $q = \infty$: $\ ^{\infty}D(\mathbf{p}) = \frac{1}{\max(\mathbf{p})}$, or Berger-Parker index

As it is obvious, when $p = \infty$, we measure diversity only based on the most common 
species, while when $p = 0$ we are counting all species equally. Hence with increasing values
of $q$ we weigh more the common species vs the rare ones in our diversity assessment. Hill numbers
have recently been proven to be the only metrics of diversity satisfying certain desirable properties
(@leinster2021).

Recently, Anne Chao and colleagues have expanded this framework to functional (and phylogenetic)
diversity, adding an extra parameter $\tau$, and calling the generalization "attribute diversity".
The daunting formula looks like:

$$
^{q, \tau}\text{AD} = \left\{\sum_{i=1}^S v_i\left(\frac{a_i(\tau)}{\bar{V}}\right)^q\right\}^{1/(1-q)}
$$

where $v_i$ are again the relative abundances of each species and $\tau$, technically an 
argument of the weighting function $a_i$ that, working as a step function,
serves as a  measure of how much resolution we apply to our distance matrix $\mathbf{D}$: any species 
separated by a distance smaller than $\tau$ are regarded as a single functional entity, 
so large values of $\tau$ capture only patterns of diversity among functionally 
(quite) different species. See @chao2019 for details.

> Note that, while the range of values of $q$ is always $(-\infty, \infty)$, the
domain of $\tau$ depends on the range of values of $\mathbf{D}$ which will change 
between different studies. 
In this study $q$ ranges between 0 and 3 and $\tau$ encompasses
the range of values of $\mathbf{D}$.

The dependence on a parameter makes of our diversity measure an entire curve (i.e. our
units of observations, the communities, are not summarized as a single number but as an
entire function varying across $q$ or $\tau$). This leads us to the use of Functional Data Analysis (FDA),
which can allow to handle response variables that consist of functions and not scalars.

When put together, AD depends on both parameters $q$ and $\tau$, generating a diversity
surface on a 3-D space. In our study we will just analyze different "slices" of this 
surface varying 

To compute Hill-Chao numbers at different parameter value combinations we used the
following code:

```{r, eval=FALSE}

# Load libraries for parallelization
library(parallel); library(foreach); library(doFuture)

# Activate cores (one for parameter combination)
cl <- parallel::makeCluster(6)
doParallel::registerDoParallel(cl)

# Set parameter grid (100 values between minimal and maximal q and tau)
{
  q_0 <- 0; q_max <- 3
  q_range <- seq(q_0, q_max, l = 1e2)
  D_traits_mat[D_traits_mat < 0] <- 1e-3
  tau_0 <- min(D_traits_mat); tau_max <- max(D_traits_mat)
  tau_range <- seq(tau_0, tau_max, l = 1e2)
  
  # This function is necessary for the a(tau) part of the AD formula:
  get_cut_D <- function(D, tau){
    D[D > tau] <- tau
    return(D)
  }
}

# Define the function
get_Hill_Chao_numbers <- function(
                                  X = as.matrix(PxS_prop_filtered),
                                  D = D_traits_mat,
                                  q, 
                                  tau = 0){
  
  # if(!all(near(rowSums(X), 1))) error("The matrix has not rowsums of 1")
  
  spp   <- apply(apply(X, 1, function(x) x > 0), 2, which)
  
  p <- vector("list", nrow(X))  
  for(i in 1:nrow(X)) p[[i]] <- X[i, `[[`(spp, i)]
  
  cut_D <- get_cut_D(D = D, tau = tau)
  
  # Hill numbers:
  get_Hill_nums <- function(q, p) sapply(q, function(q) sum(p^q)^(1/(1-q)))
  
  # Chao numbers:
  hill_chao <- vector("numeric", nrow(X))

  for(i in 1:nrow(X)){
    if(tau == 0){
      hill_chao[i] <- get_Hill_nums(q = q, p = p[[i]])
      
      # The function when q = 1 is not defined, but the lateral limits
      # are equal to the exponential of Shannon's entropy. To avoid numerical
      # instabilities, we establish a buffer of 0.001 q
      if(abs(q-1) < 1e-3) hill_chao[i] <- exp(vegan::diversity(p[[i]]))
      }
    else{
      f_tau <-  cut_D[spp[[i]], spp[[i]]]/tau
      a     <-  as.vector((1 - f_tau) %*% p[[i]])
      v     <-  p[[i]]/a
      V     <-  sum(v*a)
      
      # To address the same numerical problem as for Hill numbers:
      if(abs(q-1) < 1e-3) hill_chao[i] <- exp(sum(-v * a * log(a)))
      else hill_chao[i] <- sum(v*(a/V)^q)^(1/(1-q))
    }
  }
  
  return(hill_chao)
}

# Vectorize the operation to work for a list of parameter values
V_get_Hill_Chao_numbers <- Vectorize(get_Hill_Chao_numbers, vectorize.args = c("q", "tau"))

# Define names for the different parameter combinations (Hill prefix means we 
# vary q for a fixed tau, Chao prefix means we vary tau for a fixed q):
obj_names <- c("Hill_profiles_tmin", "Hill_profiles_tmean", "Hill_profiles_tmax",
               "Chao_profiles_0", "Chao_profiles_1", "Chao_profiles_2")

# The following is called "metaprogramming". We define objects (character strings) that
# will act as "code". Let's name the function calls:
function_names <- c("V_get_Hill_Chao_numbers(q = q_range, tau = 0)", 
                    "V_get_Hill_Chao_numbers(q = q_range, tau = mean(tau_range))", 
                    "V_get_Hill_Chao_numbers(q = q_range, tau = max(tau_range))",
                    "V_get_Hill_Chao_numbers(q = 0, tau = tau_range)",
                    "V_get_Hill_Chao_numbers(q = 1, tau = tau_range)",
                    "V_get_Hill_Chao_numbers(q = 2, tau = tau_range)"
                    )

# And now, due to the eval(str2lang(x)) combination, we can parallelize our
# operations to activate the function calls that were stored as strings:
res <- mclapply(as.list(function_names), function(x) eval(str2lang(x)), mc.cores = 6L)
```

The obtention of an entire curve of diversity led us to use FDA. This gives us an effect
of succession across the entire domain of each one of our diversity curves. Hence we can assess
the relative effect across $q$ (i.e. for common vs rare species) and across $\tau$ (i.e. 
for proximal vs distant species in the trait space). Functiona data analysis can be
made through the `mgcv` package; see @craiceanu2024.

# Rao Quadratic Entropy

The formula for Rao Quadratic Entropy is $\sum_{i,j} p_ip_jd_{ij}$, where $p_i$ and $p_j$ 
are relative abundances of different species, and $d_{ij}$ is the functional distance among them.
This can be put in a vectorized way:

$$
Q = \mathbf{p}'\mathbf{D}\mathbf{p}
$$

Where $\mathbf{p}'$ is the transpose of $\mathbf{p}$.
This type of linear algebra operations are efficiently computed in R with the `%*%` and
`crossprod` functions:

```{r, eval=FALSE}
Q <- crossprod(p, D %*% p)
```


A caveat with the use of $Q$ is that in natural communities it is strongly scale-dependent
with the number of species observed (i.e., as it is natural, species rich-communities have 
a higher range of functional distances than species poor communities in a way that changes
the expected value of their functional distances). Hence, for each community, if we 
keep the species abundance pattern $\mathbf{p}$ constant, we need to see how the observed
$Q$ compares to many other possible permutations of the $\mathbf{D}$ matrix, i.e. what is 
the *standardized effect size* (SES) of $Q$ in our community compared to the background $Q$ generated from 
different species combinations on the species pool. Due to the quadratic, hence positive
definite property of the metric, we used 4999 permutations, as the standard 999 was not
enough to enforce the Central Limit Theorem to get effect sizes following a normal distribution.

We calculated Rao's SES and p-values with the following code:

```{r, eval = FALSE}

Rao <- function(x,                     # Index of the plot (row of V) on which to compute Rao's
                V = PxS_prop_filtered, # Vegetation matrix with relative abundances
                D = D_traits_mat,      # Trait distance matrix with rownames and colnames
                reps = 5e3){           # Iterations (1+4999)
  
  p <- V[x, ] # Define vector p from the vegetation matrix
  p <- p[p > 0] # Filter out absent species
  l_p <- length(p) # Get species richness
  
  if(length(p) == 1) return(p)
  
  names(p) <- colnames(V)[which(V[x, ] > 0)]
  sp_N_pool  <- ncol(V)
  sp_nms_comm  <- names(p)
  
  rao_vec <- vector("numeric", reps)
  rao_vec[1] <- crossprod(p, D[sp_nms_comm, sp_nms_comm] %*% p)
  
  if(reps > 1){
  # Make the iterations keeping the species richness and pattern of 
  # abundance (p) constant, just changing the D matrix:
    spp_draw <- replicate(reps, sample(sp_N_pool, l_p))
    for(i in 2:reps){
      rao_vec[i] <- crossprod(p, D[spp_draw[,i], spp_draw[,i]] %*% p)
    }
    # Get standardized effect sizes
    rao_ses <- (rao_vec[1] - mean(rao_vec[-1]))/sd(rao_vec[-1])
    p_value <- pnorm(rao_ses)
    if(rao_ses > 0) p_value <- 1 - p_value
    
    return(c("Rao" = rao_vec[1], "Rao_SES"=rao_ses, "p_value" = p_value))
  }
  else return(rao_vec[1])
}
```


# Ternary Diagram of Functional Diversity

Recently, @ricotta2023 have proposed a decomposition of functional diversity into
three parts, divergence (or simply "diversity"), dominance and redundancy. Dominance
tells us how much of functional similarity is due to the event of two individuals belonging
to the same species, and redundancy is a measure of how much the trait space of
communities would be affected by the removal of a random species. This framework is
complementary to the richness-regularity-divergence presented earlier, and can help us
to make further assessments into community assembly processes from diversity patterns.

The regression on variables that form part of proportions (or of any set of variables
bound to sum a constant) needs special algorithms 
(beta regression in case of 2 proportions, Dirichlet regression in case of more than 2).
Hence, instead of using our successional gradient as we did up to now, we divided it 
in 3 classes by tertiles as early, middle and late, and performed a specific ordination
technique called "logratio analysis" as `easyCODA::LRA` (see its documentation
for references). From the ordination diagram we could obtain both the differences
in centroids between the three classes and the differences in spread through the classical
vegetation ecology functions `vegan::adonis2` and `vegan::betadisper`, respectively.

>Note that the general framework for the analysis of data in variables bound to sum
a constant (1 in case of proportions) is called compositional data analysis and
is a topic of active research. The major software tools in R are the packages 
`compositions` and `easyCODA` (from the "Girona school" and the "Barcelona school", 
respectively). See @pawlowsky-glahn2011, @greenacre2021 for references.

# Hypervolumes: Measure and Structure

Hypervolumes are clearly a functional richness measure. They are based on the parts of
the trait space that are densely occupied. As our trait space incorporates many
traits, we suffer the curse of dimensionality. Instead of using the raw trait space,
we performed an ordination on the distances and found, through the `mFD::quality.fspaces`
function. As many traits incorporate redundant information, there is a trade-off between 
the number of axes used and the loss of information by dimensionality reduction from the ordination, defining
an optima in the Mean Absolute Error (MAE) with a limited number of axes conserved from
the ordination. Although in our study the optima was found with around 9 ordination axes,
for computational reasons and the limited increase in MAE we selected 6 ordination axes.

Still, 6 axes are too many to find the hypervolumes of each community: we would need
communities with a species richness (i.e. an amount of points in the trait space) that
is not found in most vegetation plots. Hence we treated again our successional gradient
as a categorical variable and assigned each of our vegetation plots to the early-middle-late
classification. Then we used the $\phi$ coefficient as a measure of fidelity of the entire
species list to the different successional classes. To calculate the hypervolumes of
each class we made a weighted sample of all species that had $\phi_{\text{class}} > 0$ as
the species pool on which to calculate 50 iterations of the hypervolumes.

Finally, to explore the hypervolume structure, we obtained the Betti numbers of the
functional space in 1 and 2 dimensions (we did not explore higher dimensions due to
computational costs). This allows us to obtain a "persistence diagram", i.e. how 
do holes appear and disappear as we expand a buffer from each of the points in the
trait space, and hence we can see how many are there and how "persistent" they are
for each of our successional classes.

Hypervolumes are calculated through the `hypervolume` package, the $\phi$ coefficient
through the `psych` package and the holes through the `TDAstats` package.

# References {.unnumbered}